### 2024-09-14 Docs/linear-reduction.tex
Proved equivalence with parentheses

The problem can be formulated as follows: we are given the following:
\benum
    \item A set $\s X$, which is our {\it universe}.
    \item A set $\s S$ of {\it operator symbols}.
    \item For every operator symbol $\s s\in\s S$, a function $f_{\s s}\colon\s X\times\s X\longto\s X$.
    \item For every operator symbol $\s s\in\s S$, a {\it priority} $\pi(\s s)\in{\bb N}$.
\eenum
We define the following grammar of {\it expressions}:
$$ {\sl expr} \ccoloneqq ({\sl expr}) \gbar {\sl expr}\ \s S\ {\sl expr} \gbar \s X $$
\def\unparen{\hbox{\tencsc unparen}}%
All operators are left-associative.
We define $\unparen(\xi,i)$ to mean that the $i$th character in $\xi$ is not within parentheses (this can be implemented easily: iterate backwards and count how many opening and closing parentheses there
are.
If there are more opening than closing, than it is inside parentheses.)
We now define the evaluator for expressions:

\def\eval{\hbox{\tencsc eval}}
\algorithm
\Function{eval}{$\xi$}
    \If{$\xi=\s x\in\s X$}
        \State\Return $\s x$
    \ElseIf{$\xi[0]=$`('}
        \State $\xi$ is of the form $(\rho)\xi'$
        \State\Return $\eval(\eval(\rho)\xi')$
    \Else
        \Comment Find the greatest index with the smallest operator.\cr
        Note that $\max_i{\rm argmin}_if(i)$ is the maximal $i$ which minimizes $f$.\EndComment
        \State $i\coloneqq\max_i{\rm argmin}_i\set{\pi(\xi[i])}[{\xi[i]\in\s S,\unparen(\xi,i)}]$
        \State $\s s\coloneqq\xi[i]$
        \Comment $\xi'=\xi[:i-1]$, $\xi''=\xi[i+1:]$ \EndComment
        \State $\xi$ is of the form $\xi'\s s\xi''$
        \State\Return $f_{\s s}(\eval(\xi'),\eval(\xi''))$
    \EndIf
\EndFunc
\ealgorithm

For our beta reducer, our type term will be $\s X$, the atomic abstract terms will be $\s S,\Lparen,\Rparen$, and printable terms will be elements of $\s X\cup\s S$.
We will define the following initial beta reducer, we ignore states and printable term outputs because they are held constant:
\blist
    \item $\s X\ \s S\longto\s S\s X\ \snd\ (u,f\to(u,f))$
    \item $\s S\s X\ \s S\s X \longto\s S\s X\ \snd\ ((u,f),(v,g)\to(f(u,v),g))$
    \item $\s S\s X\ \s X\longto\s X\ \snd\ ((u,f),v\to f(u,v))$
    \item $\s X\ \Rparen \longto \Rparen\s X\ \snd\ (u,\_\to u)$
    \item $\s S\s X\ \Rparen\s X \longto\Rparen\s X\ \snd\ ((u,f),v\to f(u,v))$
    \item $\Lparen\ \Rparen\s X \longto\s X\ \snd\ (\_,u\to u)$
\elist
\noindent The initial state is self-explanatory.
The initial priorities map $\s x\in\s X$ to $\infty$, $\s s\in\s S$ to $\pi(\s s)$, $($ to $\infty$, and $)$ to $0$.
Since the state never changes, we can assume that all tokens are given their value in the initial state from the outset.

@@

Let us define the following set of {\it reduced expressions}, as a grammar including priorities:
$$ {\sl rexpr} \ccoloneqq \s X_n\gbar \s S\s X_n\ {\sl rexpr}\gbar \s X_n\ \s S_m\ {\sl rexpr}\ (n\geq m) $$
Say that a string of internal terms $S$ is {\it closed under $\beta$-reductions} if for every string $\xi\in S$ with arbitrary priorities, $\beta\gen\xi\in S$.

@@

Note that ${\sl expr}\subseteq{\sl rexpr}$ since ${\sl expr}\ccoloneqq\s X_\infty\gbar\s S_n\ \s X_\infty\ {\sl expr}$.
So when $\beta$-reducing an expression, we will remain inside ${\sl rexpr}$.

@@

\Proof we induct on the levels of priority in $\xi$.
If $\xi=\s x_n$ then this is trivial.
Otherwise let $n$ be the lowest priority of operator in $\xi$, so $\xi$ is of the form
$$ \xi = \xi_1\s s_n^1\xi_2\s s_n^2\cdots\xi_k\s s_n^k\xi_{k+1} $$
Now notice that
$$ \xi \to \beta^*\gen{\xi_1\s s_n^1\xi_2\cdots\xi_k\s s_n^k}\xi_{k+1} $$
by lemma 4.0.2 this is equal to
$$ \xi \to \s s_n^k[\beta^*_0\gen{\xi_1\s s_n^1\cdots\s s_n^{k-1}\xi_k}]\xi_{k+1} $$
and by lemma 4.0.3 reducing this gives
$$ f_k\bigl(\beta^*_0\gen{\xi_1\s s^1_n\cdots\s s^k_n\xi_k},\beta^*_0\gen{\xi_{k+1}}\bigr) $$
Continuing inductively on $k$, we get
$$ f_k\bigl(\cdots f_2(f_1(\beta^*_0\gen{\xi_1},\beta^*_0\gen{\xi_2}),\beta^*_0\gen{\xi_3})\cdots,\beta^*_0\gen{\xi_{k+1}}\bigr) $$
By induction, this is just
$$ f_k\bigl(\cdots f_2(f_1(\eval(\xi_1),\eval(\xi_2)),\eval(\xi_3))\cdots,\eval(\xi_{k+1})\bigr) \eqnum[bsres] $$
Now notice that
$$ \eval(\xi) = f_k(\eval(\xi_1\s s_n^1\cdots\s s_n^{k-1}\xi_k),\eval(\xi_{k+1})) $$
And continuing we get that this is equal to $(\bsres)$, as required.
\qed

>>>>

The problem can be formulated as follows: we are given the following:
\benum
    \item A set $\s X$, which is our {\it universe}.
    \item A set $\s S$ of {\it operator symbols}.
    \item For every operator symbol $\s s\in\s S$, a function $f_{\s s}\colon\s X\times\s X\longto\s X$.
    \item For every operator symbol $\s s\in\s S$, a {\it priority} $\pi(\s s)\in{\bb N}$.
\eenum
We define the following grammar of {\it expressions}:
$$ {\sl expr} \ccoloneqq \s L\ {\sl expr}\ \s R \gbar {\sl expr}\ \s S\ {\sl expr} \gbar \s X $$
\def\unparen{\hbox{\tencsc unparen}}%
All operators are left-associative.
$\s L,\s R$ represent left and right parentheses respectively.
We define $\unparen(\xi,i)$ to mean that the $i$th character in $\xi$ is not within parentheses (this can be implemented easily: iterate backwards and count how many opening and closing parentheses there
are.
If there are more opening than closing, than it is inside parentheses.)
We now define the evaluator for expressions:

\def\eval{\hbox{\tencsc eval}}
\algorithm
\Function{eval}{$\xi$}
    \If{$\xi=\s x\in\s X$}
        \State\Return $\s x$
    \ElseIf{$\xi[0]=\s L$}
        \State $\xi$ is of the form $\s L\rho\s R\xi'$
        \State\Return $\eval(\eval(\rho)\xi')$
    \Else
        \Comment Find the greatest index with the smallest operator.\cr
        Note that $\max_i{\rm argmin}_if(i)$ is the maximal $i$ which minimizes $f$.\EndComment
        \State $i\coloneqq\max_i{\rm argmin}_i\set{\pi(\xi[i])}[{\xi[i]\in\s S,\unparen(\xi,i)}]$
        \State $\s s\coloneqq\xi[i]$
        \Comment $\xi'=\xi[:i-1]$, $\xi''=\xi[i+1:]$ \EndComment
        \State $\xi$ is of the form $\xi'\s s\xi''$
        \State\Return $f_{\s s}(\eval(\xi'),\eval(\xi''))$
    \EndIf
\EndFunc
\ealgorithm

For our beta reducer, our type term will be $\s X$, the atomic abstract terms will be $\s S,\s L,\s R$ ($\s L$ for left parentheses, $\s R$ for right), and printable terms will be elements of $\s X\cup\s S$.
We will define the following initial beta reducer, we ignore states and printable term outputs because they are held constant:
\blist
    \item $\s X\ \s S\longto\s S\s X\ \snd\ (u,f\to(u,f))$
    \item $\s S\s X\ \s S\s X \longto\s S\s X\ \snd\ ((u,f),(v,g)\to(f(u,v),g))$
    \item $\s S\s X\ \s X\longto\s X\ \snd\ ((u,f),v\to f(u,v))$
    \item $\s X\ \s R \longto \s R\s X\ \snd\ (u,\_\to u)$
    \item $\s S\s X\ \s R\s X \longto\s R\s X\ \snd\ ((u,f),v\to f(u,v))$
    \item $\s L\ \s R\s X \longto\s X\ \fst\ (\_,u\to u)$
\elist
\noindent The initial state is self-explanatory.
The initial priorities map $\s x\in\s X$ to $\infty$, $\s s\in\s S$ to $\pi(\s s)$, $($ to $\infty$, and $)$ to $0$.
Since the state never changes, we can assume that all tokens are given their value in the initial state from the outset.

@@

\subsection{Expressions without Parentheses}

First we will prove this result for expressions without parentheses, i.e. those of the grammar
$$ {\sl expr} \ccoloneqq \s X_\infty\gbar \s X_\infty\ \s S_n\ {\sl expr} $$
Let us define the following set of {\it reduced expressions}, as a grammar including priorities:
$$ {\sl rexpr} \ccoloneqq \s X_n\gbar \s S\s X_n\ {\sl rexpr}\gbar \s X_n\ \s S_m\ {\sl rexpr}\ (n\geq m) $$
Note that ${\sl expr}\subseteq{\sl rexpr}$.
Say that a string of internal terms $S$ is {\it closed under $\beta$-reductions} if for every string $\xi\in S$ with arbitrary priorities, $\beta\gen\xi\in S$.

@@

Since ${\sl expr}\subseteq{\sl rexpr}$, when $\beta$-reducing an expression, we will remain inside ${\sl rexpr}$.

@@

\Proof we induct on the levels of priority in $\xi$.
If $\xi=\s x_n$ then this is trivial.
Otherwise let $n$ be the lowest priority of operator in $\xi$, so $\xi$ is of the form
$$ \xi = \xi_1\s s_n^1\xi_2\s s_n^2\cdots\xi_k\s s_n^k\xi_{k+1} $$
Now notice that
$$ \xi \tos \beta^*\gen{\xi_1\s s_n^1\xi_2\cdots\xi_k\s s_n^k}\xi_{k+1} $$
by lemma 4.0.2 this is equal to
$$ \xi \tos \s s_n^k[\beta^*_0\gen{\xi_1\s s_n^1\cdots\s s_n^{k-1}\xi_k}]\xi_{k+1} $$
and by lemma 4.0.3 reducing this gives
$$ f_k\bigl(\beta^*_0\gen{\xi_1\s s^1_n\cdots\s s^k_n\xi_k},\beta^*_0\gen{\xi_{k+1}}\bigr) $$
Continuing inductively on $k$, we get
$$ f_k\bigl(\cdots f_2(f_1(\beta^*_0\gen{\xi_1},\beta^*_0\gen{\xi_2}),\beta^*_0\gen{\xi_3})\cdots,\beta^*_0\gen{\xi_{k+1}}\bigr) $$
By induction, this is just
$$ f_k\bigl(\cdots f_2(f_1(\eval(\xi_1),\eval(\xi_2)),\eval(\xi_3))\cdots,\eval(\xi_{k+1})\bigr) \eqnum[bsres] $$
Now notice that
$$ \eval(\xi) = f_k(\eval(\xi_1\s s_n^1\cdots\s s_n^{k-1}\xi_k),\eval(\xi_{k+1})) $$
And continuing we get that this is equal to $(\bsres)$, as required.
\qed

\subsection{Expressions with Parentheses}

\def\exprp{\begingroup\sl expr\endgroup_P}
\def\rexprp{\begingroup\sl rexpr\endgroup_P}
Our expressions will now be of the grammar
$$ \exprp \ccoloneqq \s X_\infty\gbar \s X_\infty\ \s S_n\ \exprp\gbar \s L_\infty\ \exprp\ \s R_0 $$
and {\it reduced expressions} will be
$$ \rexprp \ccoloneqq \s X_n\gbar \s S\s X_n\ \rexprp\gbar \s X_n\ \s S_m\ \rexprp\ (n\geq m)\gbar \s L_\infty\ \rexprp\ \s R_0\gbar \s L_\infty\ (\s S\s X_n)^*\ \s R\s X_0 $$

\lemm
    $\rexprp$ is closed under $\beta$-reductions.
\elemm

\Proof we will prove the following stronger result:

\centerline{Let $\xi\in\rexprp$ then $\beta\gen\xi\in\rexprp$ and $\beta^*\gen\xi=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$ with $n_1<\cdots<n_{k+1}$.}
\benum
    \item For $\xi=\s X_n$ this is trivial.
    \item For $\xi=\s S\s X_n\xi'$, if $\s S\s X_n$ can reduce with the first token of $\xi'$, so $\xi'=\s X_m$ or $\xi'=\s S\s X_m\xi''$, then we reduce and continue by induction.
        Otherwise $\beta\gen\xi=\s S\s X_n\beta\gen{\xi'}$ which is inductively in $\rexprp$.
        And if $\beta\gen{\xi'}$'s length is decreased, we induct.
        Otherwise $\beta^*\gen{\xi'}=\xi'$ and so $\xi'=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$, so we get that $\xi=\s S\s X_n\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$ which we already
        know from the previous section converges to the desired form.
    \item For $\xi=\s X_n\s S_m\xi'$, we have that $\beta\gen\xi=\s S\s X_m\xi'$ and so we induct.
    \item For $\xi=\s L_\infty\xi'\s R_0$, since $\s L_\infty$ cannot reduce with the first character of $\xi'$ (as it is not $\s R\s X$), $\beta\gen\xi=\s L_\infty\beta\gen{\xi'\s R_0}$.
        If $\xi'$ can be reduced, then this is just $\s L_\infty\beta\gen{\xi'}\s R_0$ and we induct.
        Otherwise, $\xi'=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$ with $n_1<\cdots<n_{k+1}$, and so this becomes $\s L_\infty\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s R\s X_0$, which is covered by the
        final rule.
    \item For $\xi=\s L_\infty\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s R\s X_0=\s L_\infty\xi'\s R\s X_0$, if $\xi'$ can be reduced then we simply induct.
        Otherwise $n_1<\cdots<n_k$ and so a $\beta$-reduction gives $\s L_\infty\s S\s X_{n_1}\cdots\s S\s X_{n_{k-1}}\s R\s X_0$, and we induct.
        The base is taken by $k=0$ in which case we get $\s L_\infty\s R\s X_0\to\s X_\infty$.
        \qed
\eenum

From this lemma, since $\beta^*\gen\xi=\s S\s X_{n_1}\cdots\s S\s X_{n_k}\s X_{n_{k+1}}$, we get that $\beta^*_0\gen\xi\in\s X$.

Further note from $(4)$ we can infer that $\s L_\infty\xi\s R_0\tos\s L_\infty\beta^*\gen\xi\s R_0$.
From the proof of this lemma we have that $\beta^*\gen\xi=\s s_{n_1}^1[\s x^1]\cdots\s s_{n_k}^k[\s x^k]\s x_{n_{k+1}}^{k+1}$, and so we can see that
$\s L_\infty\beta^*\gen\xi\s R_0\tos\beta^*_0\gen\xi_\infty$.
Thus we have proven the following lemma:

\lemm
    Let $\xi\in\rexprp$, then $\beta^*\gen{\s L_\infty\xi\s R_0}=\beta^*_0\gen\xi_\infty$.
\elemm

\lemm
    Let $\xi\in\rexprp$ and $n\leq\pi(\xi)$, then $\beta^*\gen{\xi\s s_n}=s_n[\beta^*_0\gen\xi]$.
\elemm

\Proof we know that
$$ \xi\s s_n \tos \beta^*\gen\xi\s s_n = \s s^1_{n_1}[\s x^1]\cdots\s s^k_{n_k}[\s x^k]\s x_{n_{k+1}}^{k+1}\s s_n $$
and so this just reduces to the same lemma as for expressions without parentheses.
\qed

\lemm
    Let $\xi\in\rexprp$ and $n<\pi(\xi)$, then $\beta^*_0\gen{\s s_n[\s x]\xi}=f_{\s s}(\s x,\beta^*_0\gen\xi)$.
\elemm

\Proof this too, reduces to the same proof as for expressions without parentheses.
\qed

\thrm
    For $\xi\in\exprp$, $\eval(\xi)=\beta^*_0\gen\xi$.
\ethrm

\Proof We induct on the length of $\xi$.
If $\xi=\s x$ this is trivial.
If $\xi=\s L_\infty\rho\s R_0\xi'$ for $\rho\in\exprp$ and so
$$ \xi\tos\beta^*\gen{\s L_\infty\rho\s R_0}\xi' = \beta^*_0\gen\rho_\infty\xi' = \eval(\rho)\xi' $$
where the final equality is due to induction.
Then inductively, we have that $\beta^*_0\gen\xi=\beta^*_0\gen{\eval(\rho)\xi'}=\eval(\eval(\rho)\xi')$, which is just $\eval(\s L\rho\s R\xi')=\eval(\xi)$.
And finally if $n$ is the lowest operator priority in $\xi$, then $\xi=\xi_1\s s_n^1\xi_2\cdots\xi_k\s s_n^k\xi_{k+1}$, this just reduces to the same proof for expressions without parentheses by the above
two lemmas.
\qed
